### Homework3, Stat 627
### Zhuangye Chen
```{r setup}
# Load the packages
suppressPackageStartupMessages(library(plyr))
suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(library(stringr))
```

#### Loading dirty Gapminder
```{r read in data}
gapminder1 <- read.delim("https://raw.githubusercontent.com/STAT545-UBC/STAT545-UBC.github.io/master/gapminderDataFiveYear_dirty.txt", strip.white = FALSE)
gapminder2 <- read.delim("https://raw.githubusercontent.com/STAT545-UBC/STAT545-UBC.github.io/master/gapminderDataFiveYear_dirty.txt", strip.white = TRUE, sep = "\t")
str(gapminder1)
str(gapminder2)
sapply(1:dim(gapminder1)[2], function(x) identical(gapminder1[, x], gapminder2[, x]))
```
It turns out that the first 4 columns of these two data frames are the same while the last column is not (151 levels in gapminder1 but 148 levels in gapminder2). Now let's take a closer look at it.
```{r last col}
sum(unique(gapminder2$region) %in% unique(gapminder1$region))
# This means that every level of region in gapminder2 exists in gapminder1.
# Now we'll see which levels are in gapmider1 but not gapminder2
unique(gapminder1$region)[!unique(gapminder1$region) %in% unique(gapminder2$region)]
```
Since the paramter *strip.white* is used to allow the stripping of leading and trailing white space from unquoted character fields. It's most likely that when *strip.white = FALSE*, the white space leading or trailing the above three levels is not trimmed and thus we have these 3 extra levels in gapminder1. And we will be using gapminder2 from now on.

#### Splitting or merging
The continent and country are both in the field *region*. We will split them so that each column only contains one conceptual variable.
```{r split or merge}
region_split <- str_split_fixed(as.character(gapminder2$region), pattern = "_", n = 2)
gapminder2 <- gapminder2 %>% mutate(continent = region_split[, 1], country = region_split[, 2]) %>% select(-region)
head(gapminder2)
```

#### Missing values
Run *table()* function to character column and *is.na()* function to numeric column to see if everything looks good.
```{r missing values}
# For all numeric columns
anyNA(gapminder2[, sapply(gapminder2, FUN = is.numeric)])
# For all character columns
table(gapminder2$continent)
# If there is any empty string in the field country
table(gapminder2$country == "")
# Also, we can find those countries of frequency not equal to 12
table(gapminder2$country)[table(gapminder2$country) != 12]
```
There is missing value (empty string) in field *continent* but not in *country*.

```{r fill the missing value}
gapminder2 %>% filter(continent == "")
# Canada is from Americas, add this to the continent column
gapminder2[gapminder2$continent == "", "continent"] <- "Americas"
gapminder2 %>% filter(country == "Canada")
```

#### Inconsistent capitalization and spelling
```{r spelling}
# Those countries of frequency not equal to 12
table(gapminder2$country)[table(gapminder2$country) != 12]
# Get names for these country for capitalization
names_to_change <- names(table(gapminder2$country)[table(gapminder2$country) != 12])
gapminder2$country[gapminder2$country %in% names_to_change] <- gapminder2$country[gapminder2$country %in% names_to_change] %>% gsub("\\b([a-z])", "\\U\\1", ., perl = TRUE)
# Let's see what is still there 
table(gapminder2$country)[table(gapminder2$country) != 12]
names_to_change <- names(table(gapminder2$country)[table(gapminder2$country) != 12])
gapminder2$country[gapminder2$country %in% names_to_change] <- 
gapminder2$country[gapminder2$country %in% names_to_change] %>% gsub("(Cote )(.*)", "\\1d'Ivoire", .)

table(gapminder2$country)[table(gapminder2$country) != 12]
# At this point, it seems that manually changing the name is the fastest way 
names_to_change <- names(table(gapminder2$country)[table(gapminder2$country) != 12])
gapminder2$country[gapminder2$country %in% names_to_change] <- "Congo, Dem. Rep."
table(gapminder2$country)[table(gapminder2$country) != 12]
```
Now every country is of requency 12.


#### Final check
```{r final check}
gapminder_clean <- read.delim("https://raw.githubusercontent.com/STAT545-UBC/STAT545-UBC.github.io/master/gapminderDataFiveYear.txt", colClasses = c("character", "integer", "numeric", "character", "numeric", "numeric"))
# Swap the columns in gapminder2 and compare
gapminder2 <- gapminder2[, names(gapminder_clean)]
identical(gapminder2, gapminder_clean)
```
Data cleaning completed!